Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-12-10 manuscript/manuscript.tex
update ms. and ms-notes.

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{response times, event history analysis, Bayesian multi-level regression models, experimental psychology, cognitive psychology\newline\indent Word count: X}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\raggedbottom
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows},
  pdfauthor={Sven Panis1 \& Richard Ramsey1},
  pdflang={en-EN},
  pdfkeywords={response times, event history analysis, Bayesian multi-level regression models, experimental psychology, cognitive psychology},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the time course of psychological states. As such, the approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to discriminate between different theoretical accounts of psychological and/or brain-based processes.
For example, the distributions in Example 4 show that the effect starts between 100 and 200 ms (in timebin 2) and is gone when the waiting time reaches 500 ms or more. In contrast, in Example 5, the effect starts around 300 ms and is gone by 700 ms. And in the Example 6, the effect reverses between 500 and 600 ms. What kind of theory or theories could account for such effects? Are there new auxiliary assumptions that theories need to adopt? And are there new experiments that need to be performed to test the novel predictions that follow from these analyses? As we show later using published examples, for many psychological questions, such ``temporal states'' information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes, as well as adding a relatively under-used dimension -- the passage of time -- to the theory building toolkit.

@@

We provide six different tutorials, which are written in the R programming language and publicly available on our Github and the Open Science Framework (OSF) pages, along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply EHA to their own time-to-event data sets. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as a .html file, so that it can be viewed by any web browser, and thus available to those that do not use R. Finally, the manuscript itself is written in R using the papaja package (Aust \& Barth, 2024), which makes it computationally reproducible, in terms of the underlying data and figures.

@@

In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We also perform prior predictive checks, compare models, and interpret the plots of the predicted hazard functions for the selected model, and the posterior distributions of our contrasts of interest. In Tutorial 2b we fit Bayesian multi-level regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the EHA of RT data for choice RT data.

@@

In tutorial 4, we illustrate one approach to planning how much data to collect in an experiment using EHA. We use data simulation techniques to vary sample size and trial count per condition until a certain degree of statistical power or precision is reached. {[}{[}more to come here, once we have written the tutorial{]}{]}.

In summary, even though EHA is a widely used statistical tool and there already exist many excellent reviews (e.g., Blossfeld \& Rohwer, 2002; Box-Steffensmeier, 2004; Hosmer, Lemeshow, \& May, 2011; Teachman, 1983) and tutorials (e.g., Allison, 2010; Landes, Engelhardt, \& Pelletier, 2020) on its general use-cases, we are not aware of any tutorials that are aimed at psychological time-to-event data, and which provide worked examples of the key data processing and multi-level regression modelling steps.
Therefore, our ultimate goal is twofold: first, we want to convince readers of the many benefits of using EHA when dealing with time-to-event data with a focus on psychological time-to-event data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time EHA on time-to-event data such as RT data, as well as a complementary discrete-time SAT analysis on timed accuracy data.

@@

For a comprehensive background context to EHA, we recommend several excellent textbooks (Allison, 2010; Singer \& Willett, 2003). Likewise, for a general introduction to understanding regression equations, we recommend several excellent textbooks (Gelman, Hill, \& Vehtari, 2020; Winter, 2019). Our focus here is not on providing a detailed account of the underlying regression equations, since this topic has been comprehensively covered many times before. Instead, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. As such, we only supply regression equations in part D of the supplementary material.

@@

\caption{Discrete-time hazard and survivor functions. Discrete time-to-event data were simulated for 200 trials of 1 experimental condition. While the hazard function is the vehicle for inferring the time course of cognitive processes, the survival probability S(t-1) can help to qualify or provide context to the interpretation of the hazard probability h(t). For example, the high hazard of .60 = h(t=6) is experienced only by 44 percent of the trials, as S(t=5) = .44.
Because the survivor function is a decreasing function of time, the error bars in later parts of the hazard function will always be wider and less precise compared to earlier parts.}\label{fig:plot2}
\end{figure}

@@

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. We do not cover these benefits in detail here, as these are more general topics that have been covered elsewhere in textbooks. Instead, we briefly summarise list the benefits below, and refer the reader to section F of Supplementary Materials for more detailed coverage of the benefits. A summary of the benefits are as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Hazard functions are more diagnostic than density functions when one is interested in studying the detailed shape of a RT distribution (Holden et al., 2009).
\item
  RT distributions may differ from each other in multiple ways, and hazard functions allow one to capture these differences which mean-average comparisons may conceal (Townsend, 1990).
\item
  EHA takes account of more of the data collected in a typical speeded response experiment, by virtue of not discarding right-censored observations. Trials with very long RTs are not discarded, but instead contribute to the risk set in each time bin (see below).
\item
  Hazard modeling allows one to incorporate time-varying explanatory covariates, such as heart rate, electroencephalogram (EEG) signal amplitude, gaze location, etc. (Allison, 2010). This is useful for linking physiological effects to behavioral effects when performing cognitive psychophysiology (Meyer, Osman, Irwin, \& Yantis, 1988).
\item
  EHA can help to solve the problem of model mimicry, i.e., the fact that different computational models can often predict the same mean RTs as observed in the empirical data, but not necessarily the detailed shapes of the empirical RT hazard distributions. As such, EHA can be a tool to help distinguish between competing theories of cognition and brain function.
\end{enumerate}

@@

\caption{Discrete-time hazard and conditional accuracy functions. Discrete time-to-event and conditional accuracy data were simulated for 2000 trials for each of two priming conditions (congruent and incongruent prime stimuli). Bin width equals 100 ms.}\label{fig:plot3}
\end{figure}

@@

Our lab adopts a estimation approach to multi-level regression (Kruschke \& Liddell, 2018; Winter, 2019), which is heavily influenced by the Bayesian framework as suggested by Richard McElreath (Kurz, 2023b; McElreath, 2020). We also use a ``keep it maximal'' approach to specifying varying (or random) effects (Barr, Levy, Scheepers, \& Tily, 2013). This means that wherever possible we include varying intercepts and slopes per participant
To make inferences, we use two main approaches. We compare models of different complexity, using information criteria (e.g., WAIC) and cross-validation (e.g., LOO), to evaluate out-of-sample predictive accuracy (McElreath, 2020). We also take the most complex model and evaluate key parameters of interest using point and interval estimates.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. The content of the tutorials, in terms of EHA and multi-level regression modelling, is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2020), Heiss (2021), Kurz (2023a), and Kurz (2023b).

@@

Next, we can set up the life tables and plots of the discrete-time functions h(t), S(t), ca(t), and P(t) -- see part A of the supplementary material for their definitions. To do so using a functional programming approach, one has to nest the data within participants using the group\_nest() function, and supply a user-defined censoring time and bin width to our custom function ``censor()'', as follows.

@@

\caption{Estimated discrete-time hazard, survivor, probability mass, and conditional accuracy functions for participant 6. Vertical dotted lines indicate the estimated median RTs. Error bars represent +/- 1 Standard Error of the respective proportion.}\label{fig:eha-plot}
\end{figure}

@@

In this third tutorial, we illustrate how to fit Bayesian multi-level regression models to the RT data of the masked response priming data set used in Tutorial 1a. Fitting (Bayesian or non-Bayesian) regression models to time-to-event data is important when you want to study how the shape of the hazard function depends on various predictors (Singer \& Willett, 2003).

@@

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see part C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a categorical or continuous predictor. And when you treat a variable as a categorical predictor, you can choose between reference coding and index coding. With reference coding, one defines the variable as a factor and selects one of the k categories as the reference level. Brm() will then construct k-1 indicator variables (see model M1d in Tutorial\_2a.Rmd for an example). With index coding, one constructs an index variable which contains integers that correspond to different categories (see models M0i and M1i below). As explained by McElreath (2020), the advantage of index coding is that the same prior can be assigned to each level of the index variable, so that each category has the same prior uncertainty.

@@

Because the parameter estimates are on the cloglog-hazard scale, we can ease our interpretation by plotting the expected value of the posterior predictive distribution -- the predicted hazard values -- for the average participant (Figure 6A), and for each participant in the data set (Figure 6B).

@@

\caption{Point (median) and 80/95\% credible interval summaries of the hazard estimates (expected values of the draws from the posterior predictive distributions) in each time bin for the average participant (A), and for each participant (B).}\label{fig:plot-pred-grand-effects}
\end{figure}

As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects, both for the average participant (Figure 7A; grand average marginal effect) and for each participant in the data set (Figure 7B; subject-specific average marginal effect).

@@

\caption{Point (mean) and 80/95\% credible interval summaries of estimated differences in hazard in each time bin for the average participant (A), and for each participant (B).}\label{fig:plot-grand-ame-effects}
\end{figure}

@@

If we want to study the average effect of prime type on hazard, uncontaminated by inter-individual differences, we can base our conclusion on Figure 8 and Table 4. The contrast ``congruent minus blank'' was estimated to be 0.09 hazard units in bin 6 (95\% CrI = {[}0.02, 0.17{]}), and 0.14 hazard units in bin 7 (95\% CrI = {[}0.04, 0.25{]}). For the other bins, the 95\% credible interval contained zero.
The contrast ``incongruent minus blank'' was estimated to be 0.09 hazard units in bin 6 (95\% CrI = {[}0.01, 0.21{]}), -0.19 hazard units in bin 9 (95\% CrI = {[}-0.31, -0.06{]}), -0.27 hazard units in bin 10 (95\% CrI = {[}-0.45, -0.04{]}), and -0.23 hazard units in bin 11 (95\% CrI = {[}-0.40, -0.03{]}). For the other bins, the 95\% credible interval contained zero. Note that we could also have calculated hazard ratios instead of hazard differences.

@@

If we want to focus more on inter-individual differences, we can study the subject-specific hazard functions in Figure 9. Note that three participants (1, 2, and 3) show a negative difference for the contrast ``congruent minus incongruent'' in bin (360,400{]} -- subject 2 also in bin (320,360{]}.

@@

Interestingly, all subjects show a tendency in their mean difference (congruent minus blank) to ``dip'' around that time (Figure 9). Therefore, future modeling efforts could incorporate the trial number into the model formula, in order to also study how the effects of prime type on hazard change on the long experiment-wide time scale, next to the short trial-wide time scale. In Tutorial\_2a.Rmd we provide a number of model formula that should get you going.

@@

In this fourth tutorial, we illustrate how to fit a Bayesian multi-level regression model to the timed accuracy data from the masked response priming data set used in Tutorial 1a. The general process is similar to Tutorial 2a, except that (a) we use the person-trial data set, (b) we use the logit link function, and (c) we change the priors. To keep the tutorial short, we only fitted the effects of model M1i (see Tutorial 2a) in the conditional accuracy model called M1i\_ca.

@@

\caption{Medians and 80/95\% credible intervals of the posterior distributions of the population-level parameters of model M1i\_ca.}\label{fig:plot-ca-fixed-effects}
\end{figure}

Because the parameter estimates are on the logit-ca scale, we can ease our interpretation by plotting the expected value of the posterior predictive distribution -- the predicted conditional accurcies -- for the average participant (Figure 9A), and for each participant in the data set (Figure 9B).

@@

\caption{Point (median) and 80/95\% credible interval summaries of the conditional accuracy estimates (expected values of the draws from the posterior predictive distributions) in each time bin for the average participant (A), and for each participant (B).}\label{fig:plot-pred-grand-effects-ca}
\end{figure}

As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects for the average participant (Figure 10; grand average marginal effect).

@@

\caption{Point (mean) and 80/95\% credible interval summaries of estimated differences in conditional accuracy in each time bin for the average participant.}\label{fig:plot-ca-grand-ame-effects}
\end{figure}

The point estimates and quantile intervals can be reported in a table (see Tutorial\_2b.Rmd for details).

Based on Figure 10 we see that congruent primes have a positive effect on the conditional accuracy of emitted responses in time bins (200,240{]}, (240,280{]}, and (280,320{]}, relative to the estimates in the baseline condition (blank prime; red dashed lines in Figure 14).
Incongruent primes have a negative effect on the conditional accuracy of emitted responses in those time bins, relative to the estimates in the baseline condition.

@@

The results of Simulation 1 show that if we were targeting an effect size similar to the one reported in the original study, then testing 10 participants and collecting 100 trials per condition would be enough to provide over 95\% power.
However, we could not be as confident about smaller effects, such as a hazard ratio of 50\% or 25\%.
From this simulation, we can see that somewhere between an effect size of a 50\% and 75\% reduction in hazard value, power increases to a range that most researchers would consider acceptable (i.e., \textgreater95\% power).
To probe this space a little further, we decided to run a second simulation, which varied different parameters

@@

\caption{Statistical power across data Simulation 1 (A) and Simulation 2 (B). Power was calculated as the percentage of lower-bound 95\% confidence intervals that exclude zero when the difference between prime condition is calculated (congruent - incongruent). In Simulation 1, the effect size was varied between a 25\%, 50\% and 75\% reduction in hazard value, whereas the trial count was varied between 100, 200 and 400 trials per condition (the number of participants was fixed at N=10). In Simulation 2, the effect size was varied between a 50\%, 60\% and 70\% reduction in hazard value, whereas the number of participants was varied between N=10, 15 and 20 (the number of trials per condition was fixed at 200).The dashed lines represent 80\% (red), 90\% (black) and 95\% (blue) power. Abbreviations: rep\_n = the number of trials per experimental condition; sub\_j = the number of participants per simulated experiment.}\label{fig:power}
\end{figure}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{response times, event history analysis, Bayesian multilevel regression models, experimental psychology, cognitive psychology\newline\indent Word count: 13574}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\raggedbottom
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows},
  pdfauthor={Sven Panis1 \& Richard Ramsey1},
  pdflang={en-EN},
  pdfkeywords={response times, event history analysis, Bayesian multilevel regression models, experimental psychology, cognitive psychology},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the time course of psychological states. As such, the approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to discriminate between different theoretical accounts of psychological and/or brain-based processes.
For example, the distributions in example 4 show that the effect starts between 100 and 200 ms (in timebin 2) and is gone when the waiting time reaches 500 ms or more. In contrast, in example 5, the effect starts around 300 ms and is gone by 700 ms. And in example 6, the effect reverses between 500 and 600 ms. What kind of theory or theories could account for such effects? Are there new auxiliary assumptions that theories need to adopt? And are there new experiments that need to be performed to test the novel predictions that follow from these analyses? As we show later using published examples, for many psychological questions, such ``temporal states'' information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes, as well as adding a relatively under-used dimension -- the passage of time -- to the theory building toolkit.

@@

We provide seven different tutorials, which are written in the R programming language and publicly available on our Github page (\url{https://github.com/sven-panis/} Tutorial\_Event\_History\_Analysis), along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply EHA to their own time-to-event data sets. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as a .html file, so that it can be viewed by any web browser, and thus available to those that do not use R. Finally, the manuscript itself is written in R using the papaja package (Aust \& Barth, 2024), which makes it computationally reproducible, in terms of the underlying data and figures.

@@

In Tutorial 2a, we illustrate how one can fit Bayesian multilevel regression models to RT data using the R package brms. We also perform prior predictive checks, compare models, and interpret the plots of the predicted hazard functions for the selected model, and the posterior distributions of our contrasts of interest. In Tutorial 2b we fit Bayesian multilevel regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the EHA of RT data for choice RT data.

@@

In tutorial 4, we illustrate one approach to planning how much data to collect in an experiment using EHA. We use data simulation techniques to vary sample size and trial count per condition until a certain degree of statistical power or precision is reached.

In summary, even though EHA is a widely used statistical tool and there already exist many excellent reviews (e.g., Blossfeld \& Rohwer, 2002; Box-Steffensmeier, 2004; Hosmer, Lemeshow, \& May, 2011; Teachman, 1983) and tutorials (e.g., Allison, 2010; Landes, Engelhardt, \& Pelletier, 2020) on its general use-cases, we are not aware of any tutorials that are aimed at psychological time-to-event data, and which provide worked examples of the key data processing and multilevel regression modelling steps.
Therefore, our ultimate goal is twofold: first, we want to convince readers of the many benefits of using EHA when dealing with time-to-event data with a focus on psychological time-to-event data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time EHA on time-to-event data such as RT data, as well as a complementary discrete-time SAT analysis on timed accuracy data.

@@

For a comprehensive background context to EHA, we recommend several excellent textbooks (Allison, 2010; Singer \& Willett, 2003). Likewise, for a general introduction to understanding regression equations, we recommend several excellent textbooks (Gelman, Hill, \& Vehtari, 2020; Winter, 2019). Our focus here is not on providing a detailed account of the underlying regression equations, since this topic has been comprehensively covered many times before. Instead, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. As such, we only supply regression equations in section D of the supplementary material.

@@

\caption{Discrete-time hazard and survivor functions. Discrete time-to-event data were simulated for 200 trials of 1 experimental condition. Error bars represent +/- 1 standard error of the respective proportion. While the hazard function is the vehicle for inferring the time course of cognitive processes, the survival probability S(t-1) can help to qualify or provide context to the interpretation of the hazard probability h(t). For example, the high hazard of .60 = h(t=6) is experienced only by 44 percent of the trials, as S(t=5) = .44.
Because the survivor function is a decreasing function of time, the error bars in later parts of the hazard function will always be wider and less precise compared to earlier parts.}\label{fig:plot2}
\end{figure}

@@

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. We do not cover these benefits in detail here, as these are more general topics that have been covered elsewhere in textbooks. Instead, we briefly list the benefits below, and refer the reader to section F of the supplementary material for more detailed coverage of the benefits. The benefits include:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Hazard functions are more diagnostic than density functions when one is interested in studying the detailed shape of a RT distribution (Holden et al., 2009).
\item
  RT distributions may differ from each other in multiple ways, and hazard functions allow one to capture these differences which mean-average comparisons may conceal (Townsend, 1990).
\item
  EHA takes account of more of the data collected in a typical speeded response experiment, by virtue of not discarding right-censored observations. Trials with very long RTs are not discarded, but instead contribute to the risk set in each time bin (see section 4.1.2 below).
\item
  Hazard modeling allows one to incorporate time-varying explanatory covariates, such as heart rate, electroencephalogram (EEG) signal amplitude, gaze location, etc. (Allison, 2010). This is useful for linking physiological effects to behavioral effects when performing cognitive psychophysiology (Meyer, Osman, Irwin, \& Yantis, 1988).
\item
  EHA can help to solve the problem of model mimicry, i.e., the fact that different computational models can often predict the same mean RTs as observed in the empirical data, but not necessarily the detailed shapes of the empirical RT hazard distributions. As such, EHA can be a tool to help distinguish between competing theories of cognition and brain function.
\end{enumerate}

@@

\caption{Discrete-time hazard and conditional accuracy functions. Discrete time-to-event and conditional accuracy data were simulated for 2000 trials for each of two priming conditions (congruent and incongruent prime stimuli). Error bars represent +/- 1 standard error of the respective proportion. Bin width equals 100 ms.}\label{fig:plot3}
\end{figure}

@@

Our lab adopts a estimation approach to multilevel regression (Kruschke \& Liddell, 2018; Winter, 2019), which is heavily influenced by the Bayesian framework as suggested by Richard McElreath (Kurz, 2023b; McElreath, 2020). We also use a ``keep it maximal'' approach to specifying varying (or random) effects (Barr, Levy, Scheepers, \& Tily, 2013). This means that wherever possible we include varying intercepts and slopes per participant
To make inferences, we use two main approaches. We compare models of different complexity, using information criteria (e.g., WAIC) and cross-validation (e.g., LOO), to evaluate out-of-sample predictive accuracy (McElreath, 2020). We also take the most complex model and evaluate key parameters of interest using point and interval estimates.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2020), Heiss (2021), Kurz (2023a), and Kurz (2023b).

@@

Next, we can set up the life tables and plots of the discrete-time functions h(t), S(t), ca(t), and P(t) -- see section A of the supplementary material for their definitions. To do so using a functional programming approach, one has to nest the data within participants using the group\_nest() function, and supply a user-defined censoring time and bin width to our custom function ``censor()'', as follows.

@@

\caption{Estimated discrete-time hazard, survivor, probability mass, and conditional accuracy functions for participant 6. Vertical dotted lines indicate the estimated median RTs. Error bars represent +/- 1 standard error of the respective proportion.}\label{fig:eha-plot}
\end{figure}

@@

In this third tutorial, we illustrate how to fit Bayesian multilevel regression models to the RT data of the masked response priming data set used in Tutorial 1a. Fitting (Bayesian or non-Bayesian) regression models to time-to-event data is important when you want to study how the shape of the hazard function depends on various predictors (Singer \& Willett, 2003).

@@

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see section C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a categorical or continuous predictor. And when you treat a variable as a categorical predictor, you can choose between reference coding and index coding. With reference coding, one defines the variable as a factor and selects one of the k categories as the reference level. Brm() will then construct k-1 indicator variables (see model M1d in Tutorial\_2a.Rmd for an example). With index coding, one constructs an index variable which contains integers that correspond to different categories (see models M0i and M1i below). As explained by McElreath (2020), the advantage of index coding is that the same prior can be assigned to each level of the index variable, so that each category has the same prior uncertainty.

@@

Because the parameter estimates are on the cloglog-hazard scale, we can ease our interpretation by plotting the expected value of the posterior predictive distribution -- the predicted hazard values -- for the grand average (Figure 6A), and for each participant in the data set (Figure 6B).

@@

\caption{Point (median) and 80/95\% credible interval summaries of the hazard estimates (expected values of the draws from the posterior predictive distributions) in each time bin for the grand average (A), and for each participant (B).}\label{fig:plot-pred-grand-effects}
\end{figure}

As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects, both for the grand average (Figure 7A; grand average marginal effect) and for each participant in the data set (Figure 7B; subject-specific average marginal effect).

@@

\caption{Point (mean) and 80/95\% credible interval summaries of estimated differences in hazard in each time bin for the grand average (A), and for each participant (B).}\label{fig:plot-grand-ame-effects}
\end{figure}

@@

If we want to study the average effect of prime type on hazard, uncontaminated by inter-individual differences, we can base our conclusion on Figure 7A. The contrast ``congruent minus blank'' was estimated to be 0.09 hazard units in bin (200,240{]} (95\% CrI = {[}0.02, 0.17{]}), and 0.14 hazard units in bin (240,280{]}) (95\% CrI = {[}0.04, 0.25{]}). For the other bins, the 95\% credible interval contained zero.
The contrast ``incongruent minus blank'' was estimated to be 0.09 hazard units in bin (200,240{]} (95\% CrI = {[}0.01, 0.21{]}), -0.19 hazard units in bin (320,360{]} (95\% CrI = {[}-0.31, -0.06{]}), -0.27 hazard units in bin (360,400{]} (95\% CrI = {[}-0.45, -0.04{]}), and -0.23 hazard units in bin (400,440{]} (95\% CrI = {[}-0.40, -0.03{]}). For the other bins, the 95\% credible interval contained zero.

@@

If we want to focus more on inter-individual differences, we can study the subject-specific hazard functions in Figure 7B. Note that three participants (1, 2, and 3) show a negative difference for the contrast ``congruent minus incongruent'' in bin (360,400{]} -- subject 2 also in bin (320,360{]}.

@@

Interestingly, all subjects show a tendency in their mean difference (congruent minus blank) to ``dip'' around that time (Figure 7B). Therefore, future modeling efforts could incorporate the trial number into the model formula, in order to also study how the effects of prime type on hazard change on the long experiment-wide time scale, next to the short trial-wide time scale. In Tutorial\_2a.Rmd we provide a number of model formulae that should get you going.

@@

In this fourth tutorial, we illustrate how to fit a Bayesian multilevel regression model to the timed accuracy data from the masked response priming data set used in Tutorial 1a. The general process is similar to Tutorial 2a, except that (a) we use the person-trial data set, (b) we use the logit link function, and (c) we change the priors. To keep the tutorial short, we only fitted the effects of model M1i (see Tutorial 2a) in the conditional accuracy model called M1i\_ca.

@@

\caption{Medians and 80/95\% credible intervals of the posterior distributions of the population-level parameters of model M1i\_ca. ca = conditional accuracy.}\label{fig:plot-ca-fixed-effects}
\end{figure}

Because the parameter estimates are on the logit-ca scale, we can ease our interpretation by plotting the expected value of the posterior predictive distribution -- the predicted conditional accurcies -- for the grand average (Figure 9A), and for each participant in the data set (Figure 9B).

@@

\caption{Point (median) and 80/95\% credible interval summaries of the conditional accuracy estimates (expected values of the draws from the posterior predictive distributions) in each time bin for the grand average (A), and for each participant (B).}\label{fig:plot-pred-grand-effects-ca}
\end{figure}

As we are actually interested in the effects of congruent and incongruent primes, relative to the blank prime condition, we can construct two contrasts (congruent-blank, incongruent-blank), and plot the posterior distributions of these contrast effects for the grand average (Figure 10; grand average marginal effect).

@@

\caption{Point (mean) and 80/95\% credible interval summaries of estimated differences in conditional accuracy in each time bin for the grand average.}\label{fig:plot-ca-grand-ame-effects}
\end{figure}

Based on Figure 10 we see that congruent primes have a positive effect on the conditional accuracy of emitted responses in time bins (200,240{]}, (240,280{]}, and (280,320{]}, relative to the estimates in the baseline condition (blank prime; red dashed lines in Figure 10).
Incongruent primes have a negative effect on the conditional accuracy of emitted responses in those time bins, relative to the estimates in the baseline condition.

@@

The results of Simulation 1 show that if we were targeting an effect size similar to the one reported in the original study, then testing 10 participants and collecting 100 trials per condition would be enough to provide over 95\% power.
However, we could not be as confident about smaller effects, such as a hazard ratio of 50\% or 25\%.
From this simulation, we can see that somewhere between an effect size of a 50\% and 75\% reduction in hazard value, power increases to a range that most researchers would consider acceptable (i.e., \textgreater95\% power).
To probe this space a little further, we decided to run a second simulation, which varied different parameters.

@@

\caption{Statistical power across data Simulation 1 (A) and Simulation 2 (B). Power was calculated as the percentage of lower-bound 95\% confidence intervals that exclude zero when the difference between prime condition is calculated (congruent - incongruent). In Simulation 1, the effect size was varied between a 25\%, 50\% and 75\% reduction in hazard value, whereas the trial count was varied between 100, 200 and 400 trials per condition (the number of participants was fixed at N=10). In Simulation 2, the effect size was varied between a 50\%, 60\% and 70\% reduction in hazard value, whereas the number of participants was varied between N=10, 15 and 20 (the number of trials per condition was fixed at 200). The dashed lines represent 80\% (red), 90\% (black) and 95\% (blue) power. Abbreviations: rep\_n = the number of trials per experimental condition; subj\_n = the number of participants per simulated experiment.}\label{fig:power}
\end{figure}
